---
title: "Fig2: selected analyses and plots"
output: html_document
---

Selected code is included directly below. This notebook covers only the analyses listed in its sections. See the repository README for omitted material.

```{r setup, include=FALSE}
REPO <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."), winslash = "/", mustWork = TRUE)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE, error = FALSE)
knitr::opts_knit$set(root.dir = REPO)
dir.create(file.path(REPO, "Fig2/output"), recursive = TRUE, showWarnings = FALSE)
```

The first section includes the original permutation-based analysis and may take substantially longer than the plotting section. Each section can also be executed separately.

## Taxonomic structure and rarefaction statistics

```{r analyze-taxonomic-structure}
dir.create(file.path(REPO, "Fig2/output"), recursive = TRUE, showWarnings = FALSE)


# Portability-only adapter added during code collection (2026-09-13).


# Result 2 statistical support: taxonomy-associated ARG carriage and composition
# All primary analyses use the Rep99 genome set. The script writes tabular
# evidence only; it does not add a statistics panel to Fig. 2.

required_packages <- c("readr", "dplyr", "tidyr", "tibble", "broom", "emmeans", "vegan")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop("Missing R packages: ", paste(missing_packages, collapse = ", "), call. = FALSE)
}

library(readr)
library(dplyr)
library(tidyr)
library(tibble)

project_dir <- REPO
data_dir <- file.path(REPO, "Fig2/output")
s3_path <- file.path(REPO, "shared/data/ARG_annotations.tsv")
s4_path <- file.path(REPO, "shared/data/genome_ARG_summary.tsv")
out_dir <- file.path(REPO, "Fig2/output")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

min_rep99_genomes <- 50L
permutations_n <- 9999L
rarefaction_n <- 50L
rarefaction_iterations <- 1000L
set.seed(20260620)

s3 <- readr::read_tsv(s3_path, show_col_types = FALSE)
s4 <- readr::read_tsv(s4_path, show_col_types = FALSE)

rep99 <- s4 %>%
  filter(genome_set == "Rep99") %>%
  transmute(
    genome_id,
    genus = if_else(is.na(genus) | genus == "", "Unclassified", genus),
    source_type = if_else(is.na(source_type) | source_type == "", "other", source_type),
    ARG_positive = ARG_positive == "yes",
    ARG_hit_count = as.integer(ARG_hit_count)
  )

rep99_hits <- s3 %>%
  filter(genome_set == "Rep99" | is_rep99 == "yes") %>%
  transmute(
    genome_id,
    genus = if_else(is.na(genus) | genus == "", "Unclassified", genus),
    ARG_gene,
    ARG_class
  )

stopifnot(nrow(rep99) == 4808L, nrow(rep99_hits) == 1424L)

major_genera <- rep99 %>%
  count(genus, name = "Rep99_genomes") %>%
  filter(Rep99_genomes >= min_rep99_genomes) %>%
  arrange(desc(Rep99_genomes))

model_data <- rep99 %>%
  semi_join(major_genera, by = "genus") %>%
  mutate(
    genus = factor(genus),
    source_type = factor(source_type),
    ARG_positive_numeric = as.integer(ARG_positive)
  )

# Genus effect on ARG-positive status after adjustment for source category.
glm_full <- glm(
  ARG_positive_numeric ~ genus + source_type,
  data = model_data,
  family = binomial()
)
glm_source_only <- update(glm_full, . ~ source_type)
glm_genus_only <- update(glm_full, . ~ genus)

genus_lrt <- anova(glm_source_only, glm_full, test = "LRT") %>%
  broom::tidy() %>%
  slice_tail(n = 1) %>%
  transmute(
    test = "Genus effect adjusted for source_type",
    df = df,
    deviance = deviance,
    chi_square = deviance,
    p_value = p.value,
    n_genomes = nrow(model_data),
    n_genera = nlevels(model_data$genus)
  )

source_lrt <- anova(glm_genus_only, glm_full, test = "LRT") %>%
  broom::tidy() %>%
  slice_tail(n = 1) %>%
  transmute(
    test = "Source_type effect adjusted for genus",
    df = df,
    deviance = deviance,
    chi_square = deviance,
    p_value = p.value,
    n_genomes = nrow(model_data),
    n_genera = nlevels(model_data$genus)
  )

glm_tests <- bind_rows(genus_lrt, source_lrt)

adjusted_genus <- emmeans::emmeans(glm_full, ~ genus, type = "response") %>%
  as.data.frame() %>%
  as_tibble() %>%
  rename(
    adjusted_probability = prob,
    asymptotic_SE = SE,
    ci_low = asymp.LCL,
    ci_high = asymp.UCL
  ) %>%
  left_join(
    model_data %>%
      group_by(genus) %>%
      summarise(
        observed_genomes = n(),
        observed_ARG_positive = sum(ARG_positive_numeric),
        .groups = "drop"
      ),
    by = "genus"
  ) %>%
  mutate(
    separation_flag = observed_ARG_positive == 0 | observed_ARG_positive == observed_genomes,
    ci_low = if_else(separation_flag, NA_real_, ci_low),
    ci_high = if_else(separation_flag, NA_real_, ci_high),
    note = if_else(
      separation_flag,
      "Complete separation: adjusted interval suppressed; use observed count and overall LRT",
      ""
    )
  )

genus_pairwise <- emmeans::contrast(
  emmeans::emmeans(glm_full, ~ genus),
  method = "pairwise",
  adjust = "BH"
) %>%
  as.data.frame() %>%
  as_tibble() %>%
  rename(p_value_BH = p.value)

# ARG gene composition among ARG-positive genomes from major genera.
composition_meta <- model_data %>%
  filter(ARG_positive) %>%
  select(genome_id, genus, source_type) %>%
  arrange(genome_id)

composition_long <- rep99_hits %>%
  semi_join(composition_meta, by = "genome_id") %>%
  distinct(genome_id, ARG_gene) %>%
  mutate(present = 1L)

composition_matrix <- composition_long %>%
  tidyr::pivot_wider(names_from = ARG_gene, values_from = present, values_fill = 0L) %>%
  right_join(composition_meta %>% select(genome_id), by = "genome_id") %>%
  arrange(match(genome_id, composition_meta$genome_id))

stopifnot(identical(composition_matrix$genome_id, composition_meta$genome_id))
composition_numeric <- composition_matrix %>%
  select(-genome_id) %>%
  as.data.frame()
rownames(composition_numeric) <- composition_meta$genome_id

jaccard <- vegan::vegdist(composition_numeric, method = "jaccard", binary = TRUE)
permanova <- vegan::adonis2(
  jaccard ~ genus + source_type,
  data = composition_meta,
  permutations = permutations_n,
  by = "margin"
) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  as_tibble() %>%
  rename(
    df = Df,
    sum_of_squares = SumOfSqs,
    R2 = R2,
    F_statistic = F,
    p_value = `Pr(>F)`
  ) %>%
  mutate(
    permutations = permutations_n,
    n_ARG_positive_genomes = nrow(composition_meta)
  )

genus_dispersion <- vegan::betadisper(
  jaccard,
  composition_meta$genus,
  add = "lingoes"
)
dispersion_test <- vegan::permutest(genus_dispersion, permutations = permutations_n)$tab %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  as_tibble() %>%
  rename(
    df = Df,
    sum_of_squares = `Sum Sq`,
    mean_square = `Mean Sq`,
    F_statistic = F,
    p_value = `Pr(>F)`
  ) %>%
  mutate(
    grouping = "genus",
    distance_correction = "Lingoes",
    permutations = permutations_n,
    n_ARG_positive_genomes = nrow(composition_meta)
  )

# Equal-depth sensitivity analysis for observed ARG gene richness.
hits_by_genome <- split(rep99_hits$ARG_gene, rep99_hits$genome_id)
major_genome_ids <- split(model_data$genome_id, model_data$genus)

rarefaction_draws <- lapply(names(major_genome_ids), function(g) {
  ids <- major_genome_ids[[g]]
  if (length(ids) < rarefaction_n) return(NULL)
  richness <- replicate(rarefaction_iterations, {
    sampled_ids <- sample(ids, rarefaction_n, replace = FALSE)
    length(unique(unlist(hits_by_genome[sampled_ids], use.names = FALSE)))
  })
  tibble(
    genus = g,
    draw = seq_len(rarefaction_iterations),
    rarefied_ARG_gene_richness = richness
  )
}) %>% bind_rows()

rarefaction_summary <- rarefaction_draws %>%
  group_by(genus) %>%
  summarise(
    genomes_per_draw = rarefaction_n,
    iterations = rarefaction_iterations,
    median_rarefied_richness = median(rarefied_ARG_gene_richness),
    mean_rarefied_richness = mean(rarefied_ARG_gene_richness),
    ci_low = unname(quantile(rarefied_ARG_gene_richness, 0.025)),
    ci_high = unname(quantile(rarefied_ARG_gene_richness, 0.975)),
    .groups = "drop"
  ) %>%
  left_join(major_genera, by = "genus") %>%
  arrange(desc(median_rarefied_richness), desc(mean_rarefied_richness))

analysis_manifest <- tibble(
  item = c(
    "Rep99 genomes", "Rep99 ARG hits", "Major-genus threshold",
    "Major genera", "Genomes in logistic model",
    "ARG-positive genomes in composition model", "PERMANOVA permutations",
    "Rarefaction genomes per draw", "Rarefaction iterations"
  ),
  value = as.character(c(
    nrow(rep99), nrow(rep99_hits), min_rep99_genomes,
    nrow(major_genera), nrow(model_data), nrow(composition_meta),
    permutations_n, rarefaction_n, rarefaction_iterations
  ))
)

readr::write_tsv(analysis_manifest, file.path(out_dir, "Result2_analysis_manifest.tsv"))
readr::write_tsv(glm_tests, file.path(out_dir, "Result2_logistic_LRT.tsv"))
readr::write_tsv(adjusted_genus, file.path(out_dir, "Result2_genus_adjusted_probabilities.tsv"))
readr::write_tsv(genus_pairwise, file.path(out_dir, "Result2_genus_pairwise_BH.tsv"))
readr::write_tsv(permanova, file.path(out_dir, "Result2_ARG_composition_PERMANOVA.tsv"))
readr::write_tsv(dispersion_test, file.path(out_dir, "Result2_ARG_composition_betadisper.tsv"))
readr::write_tsv(rarefaction_summary, file.path(out_dir, "Result2_rarefied_ARG_richness.tsv"))

message("Result 2 statistics written to: ", out_dir)
message("Major genera: ", nrow(major_genera), "; logistic n: ", nrow(model_data))
message("Composition n: ", nrow(composition_meta), "; ARG genes: ", ncol(composition_numeric))
```

## Fig2A — ARG prevalence by genus

```{r plot-Fig2A-genus-prevalence}
dir.create(file.path(REPO, "Fig2/output"), recursive = TRUE, showWarnings = FALSE)


# Portability-only adapter added during code collection (2026-09-13).


# Revised Fig. 2A — forest plot of genus-level ARG-positive prevalence.
# This alternative is written to Fig2/revised and does not overwrite the
# original Fig. 2A files.

required_packages <- c("readr", "dplyr", "forcats", "ggplot2", "svglite", "ragg", "scales")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop("Missing R packages: ", paste(missing_packages, collapse = ", "), call. = FALSE)
}

library(readr)
library(dplyr)
library(forcats)
library(ggplot2)

project_dir <- REPO
s4_path <- file.path(REPO, "shared/data/genome_ARG_summary.tsv")
revised_dir <- file.path(REPO, "Fig2/output")
source_data_dir <- file.path(REPO, "Fig2/output")
export_dir <- file.path(REPO, "Fig2/output")
dir.create(source_data_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(export_dir, recursive = TRUE, showWarnings = FALSE)

min_rep99_genomes <- 50L
warm <- "#B48770"
slate <- "#3E5870"
ci_colour <- "#93A7B9"
grid_colour <- "#E8EDF1"

s4 <- readr::read_tsv(s4_path, show_col_types = FALSE)
rep99 <- s4 %>%
  filter(genome_set == "Rep99") %>%
  mutate(
    genus = if_else(is.na(genus) | genus == "", "Unclassified", genus),
    ARG_positive_bool = ARG_positive == "yes"
  )

if (nrow(rep99) != 4808L) {
  stop("Rep99 genome count is ", nrow(rep99), "; expected 4808.", call. = FALSE)
}

plot_data <- rep99 %>%
  group_by(genus) %>%
  summarise(
    Rep99_genomes = n(),
    ARG_positive_genomes = sum(ARG_positive_bool),
    ARG_positive_rate_pct = 100 * ARG_positive_genomes / Rep99_genomes,
    .groups = "drop"
  ) %>%
  filter(Rep99_genomes >= min_rep99_genomes) %>%
  rowwise() %>%
  mutate(
    ci_low_pct = 100 * stats::binom.test(ARG_positive_genomes, Rep99_genomes)$conf.int[1],
    ci_high_pct = 100 * stats::binom.test(ARG_positive_genomes, Rep99_genomes)$conf.int[2]
  ) %>%
  ungroup() %>%
  arrange(desc(ARG_positive_rate_pct)) %>%
  mutate(
    rank = row_number(),
    highlight = if_else(rank <= 3, "Top three observed rates", "Other major genera"),
    fraction_label = paste0(
      scales::comma(ARG_positive_genomes, accuracy = 1), "/",
      scales::comma(Rep99_genomes, accuracy = 1)
    ),
    rate_label = sprintf("%.1f%%", ARG_positive_rate_pct),
    genus = factor(genus, levels = rev(genus))
  )

if (nrow(plot_data) != 14L) {
  stop("Major-genus count is ", nrow(plot_data), "; expected 14.", call. = FALSE)
}

readr::write_tsv(
  plot_data %>% mutate(genus = as.character(genus)),
  file.path(source_data_dir, "Fig2A_forest_revised_source_data.tsv")
)

fig2a_forest <- ggplot(plot_data, aes(y = genus, x = ARG_positive_rate_pct)) +
  geom_vline(
    xintercept = c(10, 20, 30),
    colour = grid_colour,
    linewidth = 0.3
  ) +
  geom_errorbar(
    aes(xmin = ci_low_pct, xmax = ci_high_pct),
    width = 0.18,
    linewidth = 0.55,
    colour = ci_colour
  ) +
  geom_point(
    aes(fill = highlight),
    shape = 21,
    size = 3.2,
    stroke = 0.45,
    colour = "white"
  ) +
  geom_text(
    aes(label = rate_label),
    nudge_x = 1.05,
    hjust = 0,
    size = 2.15,
    family = "Arial",
    colour = "#2E2E2E"
  ) +
  geom_text(
    aes(label = fraction_label),
    x = 35.8,
    hjust = 1,
    size = 2.05,
    family = "Arial",
    colour = "#5A5A5A"
  ) +
  scale_fill_manual(
    values = c(
      "Top three observed rates" = warm,
      "Other major genera" = slate
    ),
    guide = "none"
  ) +
  scale_x_continuous(
    limits = c(-0.6, 36.6),
    breaks = c(0, 10, 20, 30),
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0.01, 0))
  ) +
  labs(
    title = "A",
    subtitle = "ARG-positive prevalence across major genera",
    x = "ARG-positive genomes (%)",
    y = NULL,
    caption = "Exact 95% binomial CI; right labels: ARG-positive/total Rep99 genomes"
  ) +
  theme_classic(base_size = 7, base_family = "Arial") +
  theme(
    plot.title = element_text(size = 10, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 7.2, face = "bold", margin = margin(b = 5)),
    plot.caption = element_text(size = 5.4, colour = "#5A5A5A", hjust = 0, margin = margin(t = 5)),
    axis.text.y = element_text(size = 6.8, face = "italic", colour = "#2E2E2E"),
    axis.text.x = element_text(size = 6.3, colour = "#404040"),
    axis.title.x = element_text(size = 6.8),
    axis.line = element_line(linewidth = 0.4, colour = "black"),
    axis.ticks = element_line(linewidth = 0.35, colour = "black"),
    plot.margin = margin(5, 8, 5, 5)
  )

save_revised_panel <- function(plot, filename, width_mm = 105, height_mm = 105, dpi = 600) {
  width_in <- width_mm / 25.4
  height_in <- height_mm / 25.4

  svglite::svglite(paste0(filename, ".svg"), width = width_in, height = height_in)
  print(plot)
  dev.off()

  grDevices::cairo_pdf(
    paste0(filename, ".pdf"), width = width_in, height = height_in, family = "Arial"
  )
  print(plot)
  dev.off()

  ragg::agg_png(
    paste0(filename, ".png"), width = width_in, height = height_in,
    units = "in", res = 300, background = "white"
  )
  print(plot)
  dev.off()

  ragg::agg_tiff(
    paste0(filename, ".tiff"), width = width_in, height = height_in,
    units = "in", res = dpi, background = "white", compression = "lzw"
  )
  print(plot)
  dev.off()
}

out_base <- file.path(export_dir, "Fig2A_forest_revised")
save_revised_panel(fig2a_forest, out_base)

message("Revised Fig. 2A exported without overwriting originals: ", out_base)
message(
  "Top rates: ",
  paste(
    plot_data %>%
      slice_head(n = 5) %>%
      transmute(label = paste0(as.character(genus), "=", rate_label, " (", fraction_label, ")")) %>%
      pull(label),
    collapse = "; "
  )
)
```

